\ttitle

### 2015-08-26 Chapters/Introduction.tex
Paper Review - Half-way point
- Big error in SIMD - got it wrong way round and bad numbers in two chapters, multiple tables affected - fixed
- minor edit to main.tex - new package
- Other minor style and referencing changes

\subsection{Epidemiology of pancreatic cancer}
Tumours involving the head of the pancreas and the periampullary region account for a small proportion of gastrointestinal tumours. 
They may be broadly classified as benign and malignant. 
Most pancreatic neoplasia are malignant and arise from the exocrine component of the gland, the ductal epithelium.

Pancreatic ductal adenocarcinoma is the most common cancer of the pancreas. 
However, the head of the pancreas is anatomically related to several other epithelium lined structures that can also give rise to cancers. 
These include the distal common bile duct (cholangiocarcinoma), the duodenum (duodenal adenocarcinoma) and the ampulla (ampullary adenocarcinoma). 
Neuroendocrine tumours (NET) are a collection of a variety of neoplasia that originate from the endocrine portion of the pancreas. 
Other neoplasia such as intra-ductal papillary neoplasms (IPMN) as well as rare stromal tumours may also occur in this region. 
Occasionally, chronic pancreatitis may present with features similar to pancreatic cancer and can be morphologically, radiologically and histologically difficult to differentiate from cancer.

Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death. 
Only 21\% survive beyond the first year and 5-year survival is 3\% \parencite{cancerresearchuk_cancer_2014}. 
The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease \parencite{cancerresearchuk_cancer_2014,sener_pancreatic_1999}.

@@

The anatomical location of the pancreas, deep within the retroperitoneum, surrounded by numerous vital blood vessels including the coeliac trunk and its branches, the superior mesenteric artery, portal vein and superior mesenteric vein as well as proximity to other viscera such as the stomach, duodenum, transverse colon results in early involvement of these structures even by relatively small tumours. 
Moreover, symptoms are often absent in the early stages and when present are too non-specific to help with diagnosis. 
Obstructive jaundice is the most common presenting symptom and painless, obstructive jaundice in an elderly patient should always raise the suspicion of a neoplastic process in the head of the pancreas or the periampullary region. 
Other non-specific symptoms include weight loss, early satiety, vomiting, fatigue and pain in the epigastrium or the back. 
Occasionally, the first presentation of a pancreatic neoplasm may be in the form of an episode of acute pancreatitis.

@@

Initial assessment includes a thorough history, clinical examination and blood tests including full blood count, renal function tests, liver function tests and tumour markers, especially CA 19-9. 
However, definitive diagnosis requires cross-sectional imaging in the form of a contrast-enhanced computerised tomogram (CECT) of the abdomen using a pancreas-specific protocol (a modified form of the portal-venous phase). 
CECT of the pancreas when combined with CT Thorax also provides accurate information on staging of the disease with regards to metastasis and this can be supplemented by further imaging such as Positron Emission Tomography (PET-CT) or contrast-enhanced Magnetic Resonance Imaging (MRI) of the liver in specific cases. 
CECT-pancreas is also useful for assessing local resectability with accurate information regarding vascular involvement. 
Endoscopic Ultrasound (EUS) is used in selected patients to assess vascular involvement and for obtaining tissue samples for histological examination. 
In jaundiced patients, endoscopic retrograde cholangio pancreatography (ERCP) plays an important role in the alleviation of jaundice by placing stents across the obstructed bile duct, obtaining accurate visualisation of the biliary anatomy as well as obtaining brushings from within the bile duct for cytological examination. 
The role of preoperative biliary drainage is discussed in more detail in Section \ref{sec:preoperative_biliary_drainage}.

\subsection{Treatment of pancreatic cancer}
Pancreaticoduodenectomy followed by adjuvant chemotherapy offers the only chance of cure in patients with resectable pancreatic cancer who are fit enough to undergo surgery. 
In patients with unresectable disease or who are not fit to undergo surgery, palliative chemotherapy plays a limited role in prolonging survival. 
Assessment of resectability of lesions of the head of the pancreas is discussed in the next section while the assessment of physiological fitness and the impact of comorbidity are discussed in detail in section \ref{sec:comorbidity_risk_stratification} on p\pageref{sec:comorbidity_risk_stratification}.

\section{Surgical treatment of pancreatic cancer}
Pancreaticoduodenectomy remains a technically challenging and complex surgical procedure.% over a hundred years after its description. 
Excision of cancer of the head of the pancreas was first described as a two-stage operation by a German surgeon, Walther Kausch in 1909 at Augusta-Viktoria-Krankenhaus in Berlin-Schöneberg \parencite{kausch_carcinom_1912}. 
The operation was further popularised initially as a two-stage procedure by Allen Whipple \parencite{whipple_treatment_1935}, an American surgeon, before evolving into the current single stage operation by the 1950s \parencite{whipple_rationale_1941, whipple_radical_1950}.

\subsection{Resectability criteria}
\label{sec:resectability_criteria}
Resectable pancreatic cancer is defined as a tumour that does not involve the coeliac axis or the superior mesenteric artery and is not associated with distant metastatic disease. 
Tumours involving the portal vein or superior mesenteric vein are considered borderline resectable and can still be resected completely with en-bloc venous resection in selected cases \parencite{fuhrman_rationale_1996}. 
Neoadjuvant therapies such as chemotherapy with or without local radiotherapy \parencite{gillen_preoperative/neoadjuvant_2010, evans_preoperative_2008} and newer treatment modalities such as electroporation \parencite{bower_irreversible_2011} and intra-peritoneal chemotherapy are the subject of ongoing clinical research.
There is increasing evidence that some of these patients may be able to undergo potentially curative surgery after neo-adjuvant treatment.

@@

Patients received thrombo-prophylaxis on the night before surgery which was continued until discharge from hospital. 
General anaesthesia with complete muscle relaxation was used in all patients. 
Epidural analgesia was used routinely in patients during the early part of the study period while all patients in the later half of the study period received spinal diamorphine. 
Prophylactic antibiotics were administered at induction. 
The use of Octreotide, a somatostatin analogue, to reduce the incidence of postoperative pancreatic fistula is still under debate.
However, it was routinely used in all patients at our centre. 
Octreotide was administered intra-operatively (200 micrograms by subcutaneous injection at induction and repeated after 4 hours) and was continued for 5 days postoperatively (200 micrograms by subcutaneous injection administered three times a day).

A roof-top incision was made to enter the abdominal cavity. 
After assessing the peritoneal cavity and liver for absence of metastatic disease, an early assessment was made for local resectability. 
This involved complete Kocherisation of the duodenum to assess the retroperitoneum. 
Both the superior mesenteric artery and coeliac axis were assessed early for tumour involvement (‘artery-first’ approach). 
The gastrocolic omentum was divided to enter the lesser sac. 
The superior mesenteric vein was identified and a retro-pancreatic tunnel was created between the pancreatic neck and the portal vein. 
If less than half the circumference of the superior mesenteric vein or portal vein was involved, an en-bloc resection was performed with vein repair at the same time. 
The hepatoduodenal ligament was dissected after a fundus-first cholecystectomy to isolate the common bile duct which was transected after ascertaining the hepatic artery anatomy. 
The gastro-duodenal artery was divided leaving a short segment on the common hepatic artery. 
This was done to facilitate mesenteric angiography and embolisation of this vessel in case of post-pancreatectomy haemorrhage. 
The stomach (classical Whipple) or the first part of the duodenum (pylorus-preserving pancreaticoduodenectomy) were divided using a stapling device. 
The pancreatic neck was transected using Harmonic scalpel, diathermy or cold blade depending on surgeon preference. 
The jejunum was divided at the first jejunal arcade and the jejunal mesentery was divided using the Harmonic ultrasonic scalpel. 
The retroportal lamina was divided preserving the superior mesenteric artery.

Reconstruction was performed as follows: either a pancreatico-jejunostomy was fashioned using 4-0 Biosyn sutures in a two-layer duct-to-mucosa technique or a pancreatico-gastrostomy was created using 3/0 Biosyn sutures placed in a similar manner. 
Hepatico-jejunostomy was performed using interrupted 4/0 Biosyn sutures.
Gastro-jejunostomy or duodenojejunostomy (in pylorus-preserving pancreaticoduodenectomy) was performed using continuous 3/0 Polydioxanone (PDS) sutures in 2-layers. 
One or two surgical drains were placed and the abdomen was closed after ensuring haemostasis.

@@

\subsubsection{Post-pancreatectomy haemorrhage}
\label{sec:ch_intro_PPH}
Post-pancreatectomy haemorrhage is reported to occur in 1 to 8\% of patients undergoing pancreaticoduodenectomy. 
However, it accounts for 11\% to 38\% of mortality after pancreaticoduodenectomy. 
Post-pancreatectomy haemorrhage may either be intra-luminal into the gastrointestinal tract or intra-abdominal into the peritoneal/retro-peritoneal space. 
Post-pancreatectomy haemorrhage may be from a number of potential sources although bleeding from the stump of the gastroduodenal artery is the most common cause. 
Other potential sources include suture lines at the anastomoses, gastric/duodenal ulcers or diffuse gastritis, pseudoaneurysms of the gastro-duodenal, splenic or rarely the hepatic artery or rarely, haemobilia.

>>>>

\subsection{Epidemiology of pancreatic cancer}
Tumours involving the head of the pancreas and the periampullary region account for a small proportion of gastrointestinal tumours. 
They may be broadly classified as benign and malignant. 
Most pancreatic neoplasia are malignant \parencite{li_pancreatic_2004}.

Pancreatic ductal adenocarcinoma arising from the ductal epithelium of the exocrine part of the pancreas is the most common cancer of the pancreas. 
However, the head of the pancreas is anatomically related to several other epithelium lined structures that can also give rise to cancers. 
These include the distal common bile duct (cholangiocarcinoma), the duodenum (duodenal adenocarcinoma) and the ampulla (ampullary adenocarcinoma). 
Neuroendocrine tumours (NET) are a collection of a variety of neoplasia that originate from the endocrine portion of the pancreas. 
Other neoplasia such as intra-ductal papillary neoplasms (IPMN) as well as rare stromal tumours may also occur in this region. 
Occasionally, chronic pancreatitis may present with features similar to pancreatic cancer and can be morphologically, radiologically and histologically difficult to differentiate from cancer.

Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death. 
Only 21\% survive beyond the first year and 5-year survival is 3\% \parencite{cancerresearchuk_cancer_2011}. 
The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease \parencite{cancerresearchuk_cancer_2011,sener_pancreatic_1999}.

@@

The anatomical location of the pancreas, deep within the retroperitoneum, surrounded by numerous vital blood vessels including the coeliac trunk and its branches, the superior mesenteric artery, portal vein and superior mesenteric vein as well as proximity to other viscera such as the stomach, duodenum, transverse colon results in early involvement of these structures even by relatively small tumours \parencite{li_pancreatic_2004}. 
Moreover, symptoms are often absent in the early stages and when present are too non-specific to help with diagnosis. 
Obstructive jaundice is the most common presenting symptom and painless, obstructive jaundice in an elderly patient should always raise the suspicion of a neoplastic process in the head of the pancreas or the periampullary region. 
Other non-specific symptoms include weight loss, early satiety, vomiting, fatigue and pain in the epigastrium or the back. 
Occasionally, the first presentation of a pancreatic neoplasm may be in the form of an episode of acute pancreatitis.

@@

Initial assessment includes a thorough history, clinical examination and blood tests including full blood count, renal function tests, liver function tests and tumour markers, especially CA 19-9. 
However, definitive diagnosis requires cross-sectional imaging in the form of a contrast-enhanced computerised tomogram (CECT) of the abdomen using a pancreas-specific protocol (a modified form of the portal-venous phase). 
CECT of the pancreas when combined with CT Thorax also provides accurate information on staging of the disease with regards to metastasis and this can be supplemented by further imaging such as Positron Emission Tomography (PET-CT) or contrast-enhanced Magnetic Resonance Imaging (MRI) of the liver in specific cases. 
CECT-pancreas is also useful for assessing local resectability with accurate information regarding vascular involvement. 
Endoscopic Ultrasound (EUS) is used in selected patients to assess vascular involvement and for obtaining tissue samples for histological examination. 
In jaundiced patients, endoscopic retrograde cholangio pancreatography (ERCP) may play a role in the alleviation of jaundice in patients with cholangitis by placing stents across the obstructed bile duct, obtaining accurate visualisation of the biliary anatomy as well as obtaining brushings from within the bile duct for cytological examination. 
However, the role routine preoperative biliary drainage is controversial and is discussed in more detail in Section \ref{sec:preoperative_biliary_drainage}.

\subsection{Treatment of pancreatic cancer}
Pancreaticoduodenectomy (followed by adjuvant chemotherapy in some cancers) offers the only chance of cure in patients with resectable pancreatic cancer who are fit enough to undergo surgery \parencite{li_pancreatic_2004}
In patients with unresectable disease or who are not fit to undergo surgery, palliative chemotherapy plays a limited role in prolonging survival. 
Assessment of resectability of lesions of the head of the pancreas is discussed in the next section while the assessment of physiological fitness and the impact of comorbidity are discussed in detail in section \ref{sec:comorbidity_risk_stratification} on p\pageref{sec:comorbidity_risk_stratification}.

\section{Surgical treatment of pancreatic cancer}
Pancreaticoduodenectomy remains a technically challenging and complex surgical procedure. 
Excision of cancer of the head of the pancreas was first described as a two-stage operation by a German surgeon, Walther Kausch in 1909 at Augusta-Viktoria-Krankenhaus in Berlin-Schöneberg \parencite{kausch_carcinom_1912}. 
The operation was further popularised initially as a two-stage procedure by Allen Whipple \parencite{whipple_treatment_1935}, an American surgeon, before evolving into the current single stage operation by the 1950s \parencite{whipple_rationale_1941, whipple_radical_1950}.

\subsection{Resectability criteria}
\label{sec:resectability_criteria}
Resectable pancreatic cancer is defined as a tumour that does not involve the coeliac axis or the superior mesenteric artery and is not associated with distant metastatic disease \parencite{li_pancreatic_2004}. 
Tumours involving the portal vein or superior mesenteric vein are considered borderline resectable and can still be resected completely with en-bloc venous resection in selected cases \parencite{fuhrman_rationale_1996}. 
Neoadjuvant therapies such as chemotherapy with or without local radiotherapy \parencite{gillen_preoperative/neoadjuvant_2010, evans_preoperative_2008} and newer treatment modalities such as electroporation \parencite{bower_irreversible_2011} and intra-peritoneal chemotherapy \parencite{kamath_rationale_2009} are the subject of ongoing clinical research.
There is increasing evidence that some of these patients may be able to undergo potentially curative surgery after neo-adjuvant treatment \parencite{gillen_preoperative/neoadjuvant_2010}.

@@

Patients received thrombo-prophylaxis on the night before surgery which was continued until discharge from hospital. 
General anaesthesia with complete muscle relaxation was used in all patients. 
Epidural analgesia was used routinely in patients during the early part of the study period while all patients in the later half of the study period received spinal diamorphine. 
Prophylactic antibiotics were administered at induction. 
The use of Octreotide, a somatostatin analogue, to reduce the incidence of postoperative pancreatic fistula is still under debate \parencite{halloran_complications_2002, li_pancreatic_2004} .
However, it was routinely used in all patients at our centre. 
Octreotide was administered intra-operatively (200 micrograms by subcutaneous injection at induction and repeated after 4 hours) and was continued for 5 days postoperatively (200 micrograms by subcutaneous injection administered three times a day).

A roof-top incision was made to enter the abdominal cavity. 
After assessing the peritoneal cavity and liver for absence of metastatic disease, an early assessment was made for local resectability. 
This involved complete Kocherisation of the duodenum to assess the retroperitoneum. 
Both the superior mesenteric artery and coeliac axis were assessed early for tumour involvement (‘artery-first’ approach) \parencite{weitz_artery_2010}. 
The gastrocolic omentum was divided to enter the lesser sac. 
The superior mesenteric vein was identified and a retro-pancreatic tunnel was created between the pancreatic neck and the portal vein. 
If less than half the circumference of the superior mesenteric vein or portal vein was involved, an en-bloc resection was performed with vein repair at the same time. 
The hepatoduodenal ligament was dissected after a fundus-first cholecystectomy to isolate the common bile duct which was transected after ascertaining the hepatic artery anatomy. 
The gastro-duodenal artery was divided leaving a short segment on the common hepatic artery. 
This was done to facilitate mesenteric angiography and embolisation of this vessel in case of post-pancreatectomy haemorrhage. 
The stomach (classical Whipple) or the first part of the duodenum (pylorus-preserving pancreaticoduodenectomy) were divided using a stapling device. 
The pancreatic neck was transected using Harmonic\texttrademark\ scalpel, diathermy or cold blade depending on surgeon's preference. 
The jejunum was divided at the first jejunal arcade and the jejunal mesentery was divided using the Harmonic ultrasonic scalpel. 
The retroportal lamina was divided preserving the superior mesenteric artery.

Reconstruction was performed as follows: either a pancreatico-jejunostomy was fashioned using 4-0 Biosyn\texttrademark\ sutures in a two-layer duct-to-mucosa technique or a pancreatico-gastrostomy was created using 3/0 Biosyn\texttrademark\ sutures placed in a similar manner. 
Hepatico-jejunostomy was performed using interrupted 4/0 Biosyn\texttrademark\ sutures.
Gastro-jejunostomy or duodenojejunostomy (in pylorus-preserving pancreaticoduodenectomy) was performed using continuous 3/0 Polydioxanone (PDS\texttrademark) sutures in 2-layers. 
One or two surgical drains were placed and the abdomen was closed after ensuring haemostasis.

@@

\subsubsection{Post-pancreatectomy haemorrhage}
\label{sec:ch_intro_PPH}
Post-pancreatectomy haemorrhage is reported to occur in 1 to 8\% of patients undergoing pancreaticoduodenectomy. 
However, it accounts for 11\% to 38\% of mortality after pancreaticoduodenectomy \parencite{wente_postpancreatectomy_2007}. 
Post-pancreatectomy haemorrhage may either be intra-luminal into the gastrointestinal tract or intra-abdominal into the peritoneal/retro-peritoneal space. 
Post-pancreatectomy haemorrhage may be from a number of potential sources although bleeding from the stump of the gastroduodenal artery is the most common cause. 
Other potential sources include suture lines at the anastomoses, gastric/duodenal ulcers or diffuse gastritis, pseudoaneurysms of the gastro-duodenal, splenic or rarely the hepatic artery or rarely, haemobilia.

### 2015-08-26 Chapters/Pre_Post_SIRS.tex
Paper Review - Half-way point
- Big error in SIMD - got it wrong way round and bad numbers in two chapters, multiple tables affected - fixed
- minor edit to main.tex - new package
- Other minor style and referencing changes

\subsection{Patients}
Patients who underwent elective pancreaticoduodenectomy between January 2008 and July 2012 at the West of Scotland Pancreatic Unit at the Glasgow Royal Infirmary were included in this study. 
Patients who underwent only a trial dissection or palliative surgical bypass for unresectable disease during this period were excluded.

>>>>

\subsection{Patients}
Patients who underwent elective pancreaticoduodenectomy between January 2008 and December 2012 at the West of Scotland Pancreatic Unit at the Glasgow Royal Infirmary were included in this study. 
Patients who underwent only a trial dissection or palliative surgical bypass for unresectable disease during this period were excluded.
